---
title: "STAT2005: Workshop 1 - Revision Probability and R"
author: Thejana Kottawatta (22307822)
output:
  html_document:
    toc: true
    toc_depth: '5'
    df_print: paged
  html_notebook:
    theme: cosmo
    toc: true
    toc_depth: 5
    toc_float:
      collapsed: false
      smooth_scroll: true
    number_sections: false
---

## Objectives

This workshop consists of two parts:

- **Part 1:** Introduction to R and development environments
- **Part 2:** Probability Foundations for Simulation

By the end of this workshop, you should be able to:

1. Navigate and use R effectively within R Notebook and RStudio;

2. Implement basic R functions relevant to simulation tasks (e.g., variables, functions, loops, and random number generation);

3. Recall and apply key probability concepts that underpin simulation methods.

## Part 1: Introduction & review to R

**Mini tasks**

1. Add your name + student ID to the YAML at the top of this notebook using `author` tag.
2. To verify that you have a running notebook, run a code cell:

```{r}
sessionInfo()
```

This prints R version, platform, attached packages, locale, matrix products / BLAS info.


**Running Code Chunks in RStudio**

RStudio provides several convenient ways to run code chunks in an R Markdown notebook.

**Using the Run Button**

- Click the **Run** button at the top right of a code chunk to execute that chunk.  
- The output appears immediately below the chunk.

**Using Keyboard Shortcuts**

- **Ctrl + Enter** (Windows/Linux) or **Cmd + Enter** (macOS):  
  Runs the current line or selected code and moves the cursor to the next line.
- **Ctrl + Shift + Enter** (Windows/Linux) or **Cmd + Shift + Enter** (macOS):  
  Runs the entire code chunk.
- **Alt + Enter** (Windows/Linux) or **Option + Enter** (macOS):  
  Inserts a new code chunk.

**Running Multiple Chunks**

- **Run All Chunks Above**:  
  Click the small dropdown next to the Run button and choose *Run All Chunks Above*.
- **Run All Chunks Below**:  
  Same dropdown → *Run All Chunks Below*.
- **Run All**:  
  Use *Run All* to execute every chunk in the document sequentially.

**Running the Entire Document**

- Click **Knit** to run all chunks and render the document into HTML, PDF, or Word (depending on your YAML settings).

**Insert a code Chunk**

If you are using R Notebook/Markdown within the RStudio IDE, you can also use a keyboard shortcut or toolbar button to insert a chunk with the necessary R syntax automatically. 

- Keyboard Shortcut: Press `Ctrl + Alt + I` (Windows/Linux) or `Cmd + Option + I` (macOS).

### Learning R with the *swirl* Package

The **swirl** package is an interactive learning tool that teaches R programming and data science *inside your RStudio console*. It guides you through short lessons, checks your answers, and gives immediate feedback — perfect for beginners who want hands‑on practice.
In this workshop, you will install swirl, start a lesson, and complete a few interactive exercises.

**Why use swirl?**

- It teaches R from within R
- Lessons are short, structured, and beginner‑friendly
- You get instant feedback on your answers
- You can learn at your own pace
- It reinforces the skills you’ll use in simulation labs


```
install.packages("swirl")
library(swirl)
swirl() # begin interactive session
```


For STAT2005, the following swirl lessons are most relevant:

- **R Programming** (especially: Basic Building Blocks, Workspace and Files, Sequences, Vectors, Logic, Functions, lapply/sapply)  
- **Data Science: Foundations using R** (optional but helpful)


## Part 2: Probability Foundations for Simulation

Here is a collection of summary and reference sheets I prepared for STAT1005:

- [Summary of Theory](https://zamnolence.github.io/media/Statistics_Summary_of_Theory.pdf)
- [Probability Distributions Reference Sheet](https://zamnolence.github.io/media/Probability_Ref_Sheet.pdf)
- [Hypothesis Testing Reference Sheet](https://zamnolence.github.io/media/Hypothesis_Ref_Sheet.pdf)

### 2.1 Random Number Generator

Simulation begins with the ability to generate random‑looking numbers. Computers, however, are deterministic machines—they cannot produce true randomness on their own. Instead, they use **pseudorandom number generators (PRNGs)**: algorithms that produce sequences of numbers that behave like random samples.

In this section, we will:

- implement a simple Linear Congruential Generator (LCG)
- generate Uniform(0,1) samples
- compare our LCG to built‑in generator
  
This gives us the computational foundation for all later simulation work.


#### 2.1.1 LCGs

From the lecture, we learned that an LCG produces a sequence of integers using a recurrence relation:
$$
X_{n+1} = (aX_n + c) \mod m,
$$
where $X$ is the sequence of pseudo-random values and 

- the modulus $m, 0 < m$
- the multiplier $a, 0 < a < m$
- the increment $c, 0 \leq c < m$
- the seed or start values $X_0, 0 \leq X_0 < m$

Good choices of $(a,c,m)$ give long, well‑distributed sequences.
Poor choices give visible patterns—great for teaching, not great for real simulation.


##### Exercise 2.1 

2.1.1 Write a function `lcg()` that produces a sequence of integers using a recurrence relation above. Try different parameter values.


```{r}
lcg <- function(n, seed = 1, a = 1664525, c = 1013904223, m = 2^32) {
  x <- numeric(n) # create empty array n long. This is X(n) in the formula
  x[1] <- seed
  for (i in 2:n) {
    x[i] <- (a * x[i - 1] + c) %% m
  }
  u <- x / m
  return(u)
}
set.seed(123)
v < lcg(1000)
head(v) # set X(n) to 1000
```

2.1.2 Write a function `lcg_uniform()` that returns the LCG integer output scaled to Uniform(0,1).

```{r}
lcg_uniform <- function(n, seed, a, c, m, as.uniform = TRUE) {
  x <- numeric(n) # create empty array n long. This is X(n) in the formula
  x[1] <- seed
  for (i in 2:n) {
    x[i] <- (a * x[i - 1] + c) %% m
  }
  if (as.uniform) {
    return(x / m) # convert result to uniform distribution
  } else {
    return(x) # raw results
  }
}
set.seed(123)
u <- lcg_uniform(n = 5000, seed = 1, a = 1664525, c = 1013904223, m = 2^32)
head(u) # set X(n) to 1000
```


2.1.3 Plot a histogram of `lcg_uniform(10_000)`

```{r}
hist(u, breaks = 30, col = "skyblue", xlab = "Value", main = "", cex.axis = 1.8, cex.lab = 1.8)
```

2.1.4 Plot a histogram of the built in Uniform generator `runif()`

```{r}
set.seed(123)
u <- runif(1000, 0, 1) # built-in random generator 1000 points between 0 and 1.
hist(u, breaks = 30, col = "skyblue", xlab = "Value", main = "", cex.axis = 1.8, cex.lab = 1.8)
```


### 2.2 Probability Distributions (PDF/PMF/CDF)


**Discrete RVs (e.g., Bernoulli, Binomial)**
PMF gives the probability of each possible value
$$
P(X=x)
$$
**Continuous RVs (e.g., Exponential, Normal)**
- PDF describes the shape of the distribution
$$
f(x)
$$
- CDF gives cumulative probability
$$
F(x)=P(X\leq x)
$$
Simulation gives us *samples*.
Theory gives us *functions*.


#### 2.2.1 Example: Exponential(λ)

We already simulated Exponential(λ) using inverse transform.
Now we compare the simulated histogram to the theoretical PDF:
$$
f(x)=\lambda e^{-\lambda x},\quad x\geq 0
$$

```{r}
# Simulate
lam <- 2
samples <- -log(runif(5000)) / lam

# Theoretical PDF
x <- seq(0, 3, length.out = 300)
pdf <- lam * exp(-lam * x)

# Plot
hist(samples,
  breaks = 40,
  freq = FALSE,
  col = rgb(0, 0, 1, 0.3),
  border = "black",
  main = "Exponential(λ = 2): Simulation vs PDF",
  xlab = "x"
)

lines(x, pdf, col = "red", lwd = 2)
legend("topright",
  legend = c("Simulated", "Theoretical PDF"),
  col = c(rgb(0, 0, 1, 0.6), "red"),
  lwd = c(10, 2),
  bty = "n"
)
```

##### Exercise 2.2

The Normal distribution has PDF:
$$
f(x)=\frac{1}{\sqrt{2\pi }}e^{-x^2/2}
$$
Use `rnorm()` to simulate Normal(0, 1) data and `dnorm()` to compute the theoretical PDF, then create a comparison plot similar to the example above.

```{r}
# 1. Simulate Normal(0, 1) data
n_samples <- 5000 # previous question used 5000 random samples
samples_norm <- rnorm(n_samples, mean = 0, sd = 1) # use rnporm to simulate data.

# 2. Compute theoretical PDF
# We use a sequence of x values that covers the bulk of the distribution
x <- seq(-4, 4, length.out = 300) # smooth sequence of 300 points from -4 to 4.
pdf <- dnorm(x, mean = 0, sd = 1) # dnorm() to calculate theoretical pdf.

# 3. Create comparison plot
hist(samples_norm, # simulation
  breaks = 40,
  freq = FALSE,
  # changes y-axis from count to density. Sum of the areas of the bars = 1.
  col = rgb(0, 0, 1, 0.3), # transparent blue colour
  border = "black",
  main = "Normal(0, 1): Simulation vs PDF",
  xlab = "x",
  xlim = c(-4, 4)
)

# Add theoretical PDF line
lines(x, pdf, col = "red", lwd = 2)

# Add legend
legend("topright",
  legend = c("Simulated", "Theoretical PDF"),
  col = c(rgb(0, 0, 1, 0.6), "red"),
  lwd = c(10, 2),
  bty = "n"
)
```

#### 2.2.3 Example: Bernoulli(p)

For Bernoulli(p), the theoretical PMF:
$$
P(X=1)=p,\quad P(X=0)=1-p
$$

```{r}
# Parameter
p <- 0.3

# Simulate Bernoulli samples
samples_bern <- as.integer(runif(5000) < p)

# Empirical PMF
counts <- table(samples_bern)
empirical <- counts / sum(counts)

# Theoretical PMF
theoretical <- c("0" = 1 - p, "1" = p)

# Plot
barplot(
  rbind(empirical, theoretical),
  beside = TRUE,
  col = c(rgb(0, 0, 1, 0.6), rgb(1, 0, 0, 0.6)),
  names.arg = c("0", "1"),
  ylim = c(0, max(empirical, theoretical) * 1.2),
  main = "Bernoulli(p = 0.3): Empirical vs Theoretical PMF",
  ylab = "Probability"
)

legend(
  "topright",
  legend = c("Empirical", "Theoretical"),
  fill = c(rgb(0, 0, 1, 0.6), rgb(1, 0, 0, 0.6)),
  bty = "n"
)
```

### 2.3 Law of Large Numbers (LLNs)

The Law of Large Numbers (LLN) is one of the most important ideas in probability and simulation. It explains why simulation works and why empirical averages converge to theoretical expectations.
In simple terms:

> As we take more and more samples, the sample mean gets closer to the true mean.

This is why Monte Carlo methods are powerful: averages stabilise.
We’ll demonstrate LLN using simulation.


##### 2.3.1 LLNs with Binomial 

Consider coin tosses example,

```{r}
set.seed(123)
n <- 5000

x <- rbinom(5000, size = 1, prob = 0.5) # 1 = Head, 0 = Tail
running_mean <- cumsum(x) / seq_len(n)

plot(running_mean,
  type = "l", col = "blue",
  xlab = "Number of tosses (n)",
  ylab = "Running mean (proportion of heads)",
  main = "LLNs: Proportion of Heads",
  cex.main = 2.2,
  cex.lab = 1.8,
  cex.axis = 1.8,
)

abline(h = 0.5, lwd = 2) # true mean
```


##### 2.3.1 LLNs with Exp(1)


```{r}
set.seed(2026)
x <- rexp(n, rate = 1)
running_mean <- cumsum(x) / seq_len(n)

plot(running_mean,
  type = "l", col = "red",
  xlab = "Number of samples (n)",
  ylab = "Running mean",
  main = "LLNs: Exponential(1)",
  cex.main = 2.2,
  cex.lab = 1.8,
  cex.axis = 1.8,
)

abline(h = 1, lwd = 2) # true mean
```

##### Exercise 2.3

For a Uniform(0,1) random variable with $\mathbb{E}[X]=0.5$, use `runif()` to simulate 5000 samples and create a plot to demonstrate the LLNs.

```{r}
set.seed(123)
n <- 5000 #  5000 sample.
u <- runif(n, 0, 1) # built-in random generator 1000 points between 0 and 1.
running_mean <- cumsum(u) / seq_len(u) # cumulative sum / length of sequeunce.
plot(running_mean,
  xlim = c(0, n),
  ylim = c(0, 1),
  type = "l", col = "red",
  xlab = "Number of samples (n)",
  ylab = "Running mean",
  main = "E(X) = 0.5. Demonstraion of LLN nearing true mean.",
  cex.main = 2.2, # scale size of title font.
  cex.lab = 1.8, # scale font size of labels.
  cex.axis = 1.8,
)

abline(h = 0.5, lwd = 2, lty = 2) # lwd = line width, lty = line type.
```


